Clustering time series based on probability distributions across temporal granularities

### 2021-11-01 main.Rmd
Freeze data pre-preprocessing

Time series often have a somewhat skewed distribution and their ranges might vary greatly. It
is helpful to do a statistical transformation on the data to bring all of them to the same range or normalize each series. For the JS-based approaches, two data transformation techniques are utilised viz, Normal-Quantile Transform (NQT) and Robust scaling. NQT is a built-in transformation for computing $wpd$, which is the foundation of wpd-based distances.

_Robust scaling_ The normalised $i^{th}$ observation is denoted by $v_{norm} = \frac{v_t - p_{0.50}}{p_{0.75}-p_{0.25}}$, where $v_t$ is the actual value at the $t^{th}$ time point and $p_{0.25}$, $p_{0.50}$ and $p_{0.75}$ are the $25^{th}$, $50^{th}$ and $75^{th}$ percentile of the time series for the $i^{th}$ observation. $v_{norm}$ has zero mean and median, as well as a standard deviation of one, while the outliers are still there with the same relative connections to other values.

_Normal-Quantile transform_ The raw data for all observations is individually normal-quantile transformed (NQT) [@Krzysztofowicz1997-bv], so that the transformed data follows a standard normal distribution. NQT will make the skewed distributions bell-shaped. As a result, determining which raw distribution was used is difficult using the modified distribution. Also, multimodality is disguised or inverted. This, however, is not a problem for implementation of this methodology as distributions are characterised by quantiles and the order of the quantiles is reserved under NQT.

@@

@wang2020tsibble introduced the tidy "tsibble" data structure to assist temporal data exploration and modeling. To start with, the measured variable for each key variable (observation in this context) is a time-indexed sequence of values for various measurement variables at each time point. This sequence, however, could be shown in several ways. A shuffle of the raw sequence may represent hourly consumption throughout a day, a week, or a year. Cyclic granularities like hour-of-day, hour-of-week can be expressed in terms of the index set in the "tsibble" data structure. But the data structure changes while transporting from linear to cyclic scale of time as multiple observations of the measured variable would correspond to each category of the cyclic granularities.  Directly computing Jensen-Shannon distances between probability distributions can be very costly. Hence, in this paper, quantiles are chosen to characterize the probability distributions induced by the multiple observations for each category of the cyclic granularity. So, each category of a cyclic granularity corresponds to a list of numbers which is essentially few chosen quantiles of the multiple observations.

<!-- An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years.-->

>>>>

Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories. For the JS-based approaches, two data transformation techniques are utilised viz, Normal-Quantile Transform (NQT) and Robust scaling. NQT is a built-in transformation for computing $wpd$, which is the foundation of wpd-based distances.

_Robust scaling_ The normalised $i^{th}$ observation is denoted by $v_{norm} = \frac{v_t - q_{0.5}}{q_{0.75}-q_{0.25}}$, where $v_t$ is the actual value at the $t^{th}$ time point and $q_{0.25}$, $q_{0.5}$ and $q_{0.75}$ are the $25^{th}$, $50^{th}$ and $75^{th}$ percentile of the time series for the $i^{th}$ observation. $v_{norm}$ has zero mean and median, as well as a standard deviation of one, with the outliers having same relative connections to other values.

_Normal-Quantile transform_ The raw data for all observations is individually normal-quantile transformed (NQT) [@Krzysztofowicz1997-bv], so that the transformed data follows a standard normal distribution. NQT will make the skewed distributions bell-shaped. As a result, determining which raw distribution was used is difficult using the modified distribution. Multimodality is also concealed or reversed. As a result, while displaying transformed data using NQT, one must exercise caution. This, on the other hand, is a very useful strategy for making assymetric distributions symmetric, which often improves the clustering performance.

<!-- is not a problem for implementation of this methodology as distributions are characterised by quantiles and the order of the quantiles is reserved under NQT. -->

@@

We start with a "tsibble" (@wang2020tsibble) data structure with an index variable representing inherent ordering from past to present and a key variable that define observational units over time. The measured variable for observation is a time-indexed sequence of values. This sequence, however, could be shown in several ways. A shuffle of the raw sequence may represent hourly consumption throughout a day, a week, or a year. Cyclic granularities like hour-of-day, hour-of-week can be expressed in terms of the index set in the "tsibble" data structure. But the data structure changes while transporting from linear to cyclic scale of time as multiple observations are generated for each category and induces a probability distribution. Directly computing Jensen-Shannon distances between probability distributions can be very costly. Hence, in this paper, deciles are chosen to characterize the probability distributions. So, in the final data structure, each category of a cyclic granularity corresponds to a list of numbers which is essentially few chosen quantiles of the multiple observations.
